---
title: "Map Plots in R in a Tidyverse Way"
author: "Mark van der Broek"
date: "19/09/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Summary
 - I show how you can plot your own map in R using a few lines of code using a pipe-based workflow.
 - Several powerful functions of the `sf` packages are presented.

### Analysis
This week I worked on a project for which I needed to create a map plot with some statistics for selected European countries; I was unfamiliar with this kind of plots, so I searched online for possible solutions. I like the **tidyverse** workflow, so I naturally looked for any tutorials using this style. The first [hit](http://eriqande.github.io/rep-res-web/lectures/making-maps-with-R.html) was informative, but it didn't have a high resolution map for Europe. Furthermore, I like to be able to use any custom map, so I searched for ways to import a custom map.

[naturalearthdata.com](https://www.naturalearthdata.com) provides many open-source maps. I decided to select the world map with country borders on a 1:10m scale (can be found [here](https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_0_countries.zip)). 
```{r import_data, message=FALSE}
library(sf)       # For handling geospatial data
library(ggplot2)  # Plotting library
library(dplyr)    # Data manipulation in tidyverse way
library(ggthemes) # Additional themese for the ggplot2 library
library(knitr)    # Nice tables for this document

# This will create a natural-earth subfolder with the map data in the data folder.
if (!file.exists("data/natural-earth")) {
  tmp_file <- tempfile(fileext=".zip")
  download.file("https://www.naturalearthdata.com/http//www.naturalearthdata.com/download/10m/cultural/ne_10m_admin_0_countries.zip", 
                tmp_file)
  unzip(tmp_file, exdir = "data/natural-earth")
}
```

Importing these maps, however, was not straightforward to me. [These lecture slides](https://cran.r-project.org/doc/contrib/intro-spatial-rl.pdf) provides a way to import custom maps, but the syntax of the `sp` package seems very untuitive with `S4` objects for the polygons. Furthermore, the `SpatialDataFrame` objects do not support a pipe-based workflow. However, [this tutorial](https://edzer.github.io/UseR2017/) presents how the modern `sf` package can be used to manipulate, plot and import spatial data in a tidyverse manner. 

Importing our world map is as easy as
```{r create_map_data}
map_data <- st_read("data/natural-earth/", "ne_10m_admin_0_countries")
```

The `map_data` uses `data.frame`s for its features and saves the geometric features as a list in the column `geometry`. We can now easily explore the data in `map_data`, e.g.,
```{r describe}
features_map_data <- map_data %>%
  as_tibble() %>%
  select(-geometry) %>%
  head(10)

kable(features_map_data)
```

For this tutorial we want to focus on a European countries, hence we need to filter the data to only contain the european countries' info. Fortunately, the `map_data` contains a feature `CONTINTENT`, so we can easily filter out the unwanted countries.
```{r filter_countries}
europe_map_data <- map_data %>%
  select(NAME, CONTINENT, SUBREGION, POP_EST) %>%
  filter(CONTINENT == "Europe") 
```

Lets try to plot a map of European countries. New versions of `ggplot2` contain a function `geom_sf` which supports plotting `sf` objects directly, so lets try it...
```{r first_attempt_plot}
ggplot(europe_map_data) + geom_sf() +
  theme_minimal()
```

That does not seem to work... the reason is that, even though we removed the data of non European countries, we never changed the `bbox` setting of our data. The `bbox` object sets the longitude and latitude range for our plot, which is still for the whole europe. To change this we can use the `st_crop` function as
```{r crop_box}
europe_map_data <- europe_map_data %>%
  st_crop(xmin=-25, xmax=55, ymin=35, ymax=71)

ggplot(europe_map_data) + geom_sf() +
  theme_minimal()
```

If you're familiar with the `ggplot2` workflow, it is now easy to construct the aesthetic mappings like you're used to. Our `map_data` contains a feature `SUBREGION` and Europe is divided into Northern, Eastern, Southern and Western Europe. We can easily visualise this in our European map as
```{r europe_map_divide}
ggplot(europe_map_data) + geom_sf(aes(fill=SUBREGION)) +
  theme_minimal()
```

The `sf` has many in-built functions; one of these functions is `st_area` which can be used to compute the area of polygons. The population density of each country can be easily plotted by
```{r pop_density}
europe_map_data <- europe_map_data %>%
  mutate(area = as.numeric(st_area(.))) %>%
  mutate(pop_density = POP_EST / area)

ggplot(europe_map_data) + geom_sf(aes(fill=pop_density)) +
  theme_minimal() + 
  scale_fill_continuous_tableau(palette = "Green")
```

Using aggregating functions of the `tidyverse` package is also straight-forward. Lets create a similar population density plot but instead for each subregion of Europe.
```{r pop_density_sub}
subregion_data <- europe_map_data %>%
  group_by(SUBREGION) %>%
  summarise(area = sum(area), 
            pop_est = sum(POP_EST)) %>%
  ungroup() %>%
  mutate(pop_density = pop_est / area)

ggplot(subregion_data) + geom_sf(aes(fill=pop_density)) +
  theme_minimal() + 
  scale_fill_continuous_tableau(palette = "Green")
```

As a last exercise lets find the centroid for each country. 
```{r centroids_europe, warning=FALSE}
# First get all centroids of each European country
get_coordinates = function(data) {
  return_data <- data %>%
    st_geometry() %>%
    st_centroid() %>%
    st_coordinates() %>%
    as_data_frame()
}

europe_centres <- europe_map_data %>%
  group_by(NAME) %>%
  do(get_coordinates(.))

europe_map_data <- europe_map_data %>%
  left_join(europe_centres, by="NAME")
```

Actually, I only want to see the centroid of the Netherlands...
```{r centroid_netherlands}
netherlands_map_data = europe_map_data %>%
  filter(NAME == "Netherlands") %>%
  st_crop(xmin=1, xmax=10, ymin=50, ymax=55)

ggplot(netherlands_map_data) + geom_sf() +
  geom_point(aes(x=X, y=Y, colour="red")) + 
  theme_minimal()
```

### Setup
The analysis of this tutorial is performed using R version 3.5.1. To use the `st_crop` function from the `sf` package version 0.6.3 is needed. `geom_sf` also requires a recent version of `ggplot2`.
